---
title: "Tips and Tricks for Communicating Survival Analysis Results to Clinical Collaborators"
author: "Emily C. Zabor"
institute: "R/Medicine Conference"
date: "September 7, 2018"
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
library(tidyverse)
devtools::install_github("zabore/ezfun")
library(ezfun)
library(survival)
library(survminer)
library(knitr)
opts_chunk$set(echo = FALSE)
```


## The most common questions in cancer research relate to disease survival

\begin{center}
\includegraphics[scale = 0.45]{pubs}
\end{center}

## Survival analysis is a complex statistical procedure, so communication with collaborators is key

Strategies for conveying important information about survival analysis:

\begin{itemize}
  \setlength\itemsep{1em}
  \item[\checkbox] Be ready with \textbf{examples} to explain complex ideas
  \item[\checkbox] Use detailed \textbf{graphics} to convey information
  \item[\checkbox] Accompany important numbers with \textbf{interpretations}
\end{itemize}


## Example of a dataset with censored data for a clinical application

The `lung` dataset is available from the `survival` package in `R`. The data contain subjects with advanced lung cancer from the North Central Cancer Treatment Group.

Variable descriptions, from the documentation:

\small

\begin{itemize}
\setlength\itemsep{0em}
\item inst:	Institution code
\item \only<2>{\textcolor{blue}}{time:	Survival time in days}
\item \only<3>{\textcolor{blue}}{status:	censoring status 1=censored, 2=dead}
\item age: Age in years
\item sex: Male=1 Female=2
\item ph.ecog: ECOG performance score (0=good 5=dead)
\item ph.karno:	Karnofsky performance score (bad=0-good=100) rated by physician
\item pat.karno: Karnofsky performance score as rated by patient
\item meal.cal: Calories consumed at meals
\item wt.loss: Weight loss in last six months
\end{itemize}


## The Kaplan-Meier survival function forms the basis of most survival analyses

Using the `ggsurvplot` function from the `survminer` package:

```{r, message = FALSE}
fit1 <- survfit(Surv(time, status) ~ 1, data = lung)

surv_curve <- ggsurvplot(data = lung, 
                         fit = fit1,
                         xlab = "Months",
                         palette = msk_palette("main"),
                         legend = "none",
                         xscale = 30.4,
                         break.x.by = 182.4,
                         risk.table = TRUE,
                         risk.table.y.text = FALSE)

ggsave("./img/surv_curve.png", print(surv_curve), width = 6, height = 4.5)
```

\begin{center}
\includegraphics[scale = 0.6]{surv_curve}
\end{center}


## First just describe what is on the plot

```{r, message = FALSE}
surv_curve_sq <- ggsurvplot(data = lung, 
                         fit = fit1,
                         xlab = "Months",
                         palette = msk_palette("main"),
                         legend = "none",
                         xscale = 30.4,
                         break.x.by = 182.4, 
                         risk.table = TRUE,
                         risk.table.y.text = FALSE)

ggsave("./img/surv_curve_sq.png", print(surv_curve_sq), width = 4, 
       height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{surv_curve_sq}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
    \begin{itemize}
      \setlength\itemsep{1em}
  \item[] The x-axis is time and the y-axis is the survival function
  \item[] Step function where each step down represents a time at which one or more events occurred
  \item[] Censored subjects are usually denoted by tick marks
    \end{itemize}
  \end{column}
\end{columns}


## If appropriate give a basic description of the underlying calculations

\begin{columns}
  \begin{column}{0.6\linewidth}
    \includegraphics[scale = 0.6]{surv_curve_sq}
  \end{column}
  
  \begin{column}{0.4\linewidth}
The survival function is calculated at each time as the ratio of subjects who \textbf{did not} experience the event by that time to the total number of subjects \textbf{still at risk} at that time
  \end{column}
\end{columns}


## Question from collaborator: What is censoring?

In the context of a clinical trial:

\begin{center}
\includegraphics[scale = 0.5]{trial_anatomy}
\end{center}

\bigskip

\tiny{RICH JT, NEELY JG, PANIELLO RC, VOELKER CCJ, NUSSENBAUM B, WANG EW. A PRACTICAL GUIDE TO UNDERSTANDING KAPLAN-MEIER CURVES. Otolaryngology head and neck surgery: official journal of American Academy of Otolaryngology Head and Neck Surgery. 2010;143(3):331-336. doi:10.1016/j.otohns.2010.05.007.}


## Question from collaborator: Why do I need specialized methods to analyze time-to-event data?

```{r message = FALSE}
hist_futime <- ggplot(lung, aes(x = time, fill = factor(status))) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = msk_palette("contrast")[c(4, 8)], 
                    labels = c("Censored", "Dead")) +
  theme_ezbasic() +
  labs(x = "Days",
       y = "Count")

ggsave("./img/hist_futime.png", hist_futime, width = 3.5, height = 3.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.75]{hist_futime}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
    \begin{enumerate}
      \setlength\itemsep{1em}
      \item Censored subjects still provide information so must be appropriately included in the analysis
      \item The distribution of follow-up times is skewed
      \item[]
      \item[]
    \end{enumerate}
  \end{column}
\end{columns}


## Question from collaborator: Can I report the percentage of events out of the total study population?

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{surv_curve_sq}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] In the \texttt{lung} data this would lead to an estimate of survival probability at the end of the study of $1 - \frac{165}{228} = 0.28$.
    \item[] But this is \textbf{incorrect}. 
    \item[] Why?
  \end{itemize}
  \end{column}
\end{columns}


## The survival curve is lower when there is censoring during follow-up

```{r, message = FALSE}
fakedata2 <- lung %>% 
  mutate(time = ifelse(status == 2, time, 1022), 
         group = "No censoring") %>% 
  full_join(mutate(lung, group = "With censoring"))

fit3 <- survfit(Surv(time, status) ~ group, data = fakedata2)

wrong_pct <- ggsurvplot(data = fakedata2, 
                         fit = fit3,
                         xlab = "Months",
                         palette = msk_palette("main")[c(2, 1)],
                         legend = "bottom",
                         legend.title = "",
                         legend.labs = c("No censoring", 
                                         "With censoring"),
                         xscale = 30.4,
                         break.x.by = 182.4, 
                        risk.table = TRUE,
                        risk.table.y.text = FALSE)

ggsave("./img/wrong_pct.png", print(wrong_pct), width = 4.5, height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_pct}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] Imagine two studies, each with 228 subjects. 
    \item[] There are 165 deaths in each study. 
    \item[] The \textcolor{orange}{orange} study has no censoring during follow-up, the \textcolor{blue}{blue} study has subjects censored throughout (the true lung data).
  \end{itemize}
  \end{column}
\end{columns}


## When there is no censoring during follow-up, the survival estimate is straightforward

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_pct}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] In the \textcolor{orange}{orange} study everyone is followed until the last follow-up time (i.e. no censoring during follow-up). 
    \item[] The survival probability at last follow-up is \textbf{$0.28$}
    \item[] This is equivalent to an estimating the raw percent of patients still alive: $1 - \frac{165}{228} = 0.28$
  \end{itemize}
  \end{column}
\end{columns}


## When there is censoring during follow-up, a naive estimate will be incorrect

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_pct}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] The \textcolor{blue}{blue} study has $63$ subjects censored during follow-up. Still $165$ subjects died.
    \item[] The \textcolor{blue}{blue} curve is lower than the \textcolor{orange}{orange} curve, especially at the end.
    \item[] The survival probability at end of study = something $<0.28$ (actually $0.005$)
  \end{itemize}
  \end{column}
\end{columns}


## Ignoring censoring leads to an overestimate of the overall survival probability

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_pct}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  This occurs because the censored subjects only contribute information for \textbf{part} of the follow-up time, and then fall out of the risk set, thus pulling down the cumulative probability of survival.
  \end{column}
\end{columns}


## Question from collaborator: What is the 1-year survival probability?

```{r}
est_1_year <- summary(fit1, times = 365.25)$surv
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{surv_curve_sq}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] In the \texttt{lung} data the 1-year survival probability is `r round(est_1_year, 2)`
    \item[] But where does this number come from?
    \item[] And what does it mean?
  \end{itemize}
  \end{column}
\end{columns}


## We start by showing what the 1-year survival probability is on the survival curve

```{r}
x_year_est_1 <- surv_curve_sq

x_year_est_1$plot <- x_year_est_1$plot + 
  geom_segment(x = 365.25, xend = 365.25, y = -0.05, yend = 0.4092416, 
               col = msk_palette("main")[4], size = 1.5) +
  geom_segment(x = 365.25, xend = -40, y = 0.4092416, yend = 0.4092416, 
               col = msk_palette("main")[4], size = 1.5, 
               arrow = arrow(length = unit(0.2, "inches"))) 

ggsave("./img/x_year_est_1.png", print(x_year_est_1), width = 4, height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{x_year_est_1}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] The 1-year survival probability is the \textbf{probability on the y-axis} corresponding to \textbf{1-year on the x-axis}
    \item[] The \texttt{geom\_segment} function from \texttt{ggplot2} can help with this
  \end{itemize}
  \end{column}
\end{columns}


## Next we annotate the plot with text to state the probability alongside the curve

```{r}
x_year_est_2 <- x_year_est_1

x_year_est_2$plot <- x_year_est_2$plot +
  annotate("text", x = 450, y = 0.5, 
           label = glue::glue("The 1-year survival \n probability is ", 
                              "{round(est_1_year, 2)}."), 
           col = msk_palette("main")[4], hjust = 0, fontface = 2)

ggsave("./img/x_year_est_2.png", print(x_year_est_2), width = 4, height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{x_year_est_2}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
The \texttt{glue} function from the \texttt{glue} package provides an easy way to reproducibly annotate plots
  \end{column}
\end{columns}


## Additionally create reproducible text to make sure numbers are being correctly interpreted

As part of a reproducible workflow that includes writing reports in `R Markdown`, the `glue_data` function from the `glue` package can easily print reproducible text as a corresponding description:

\scriptsize

```{r echo = TRUE}
fit1 <- survival::survfit(
  survival::Surv(time, status) ~ 1, data = lung)
glue::glue_data(summary(fit1, times = 365.25), 
                "The survival probability at 1-year is ", 
                "{round(surv, 2)} ", 
                "(95% CI: ", "{round(lower, 2)} - ", "{round(upper, 2)}", 
                "),\n which represents the estimated proportion of ",
                "patients who\n survived beyond 1 year.")
```


## Finally put everything together into a report to our collaborator about the 1-year survival probability

```{r}
ggsave("./img/x_year_est_3.png", print(x_year_est_2), width = 5, height = 4.5)
```

\begin{center}
    \includegraphics[scale = 0.5]{x_year_est_3}
\end{center}

\scriptsize

```{r}
glue::glue_data(summary(fit1, times = 365.25), 
                "The survival probability at 1-year is ", 
                "{round(surv, 2)} ", 
                "(95% CI: ", "{round(lower, 2)} - ", "{round(upper, 2)}", 
                "),\n which represents the estimated proportion of ",
                "patients who\n survived beyond 1 year.")
```


## Question from collaborator: What is the median survival time?

```{r}
est_median <- summary(fit1)$table["median"]
```

\begin{columns}
  \begin{column}{0.6\linewidth}
    \includegraphics[scale = 0.6]{surv_curve_sq}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] In the \texttt{lung} data the median survival time is `r round(est_median, 1)` days
    \item[] But where does this number come from?
    \item[] And what does it mean?
  \end{itemize}
  \end{column}
\end{columns}


## We start by showing what the median survival time is on the survival curve

```{r}
median_est_1 <- surv_curve_sq

median_est_1$plot <- median_est_1$plot + 
  geom_segment(x = -45, xend = 310, y = 0.5, yend = 0.5, 
               col = msk_palette("contrast")[4], size = 1.5) +
  geom_segment(x = 310, xend = 310, y = 0.5, yend = -0.03, 
               col = msk_palette("contrast")[4], size = 1.5, 
               arrow = arrow(length = unit(0.2, "inches")))

ggsave("./img/median_est_1.png", print(median_est_1), width = 4, height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{median_est_1}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  The median survival time is the \textbf{time on the x-axis} corresponding to a \textbf{survival probability of $\textbf{0.5}$ on the y-axis}.
  \end{column}
\end{columns}


## Next we annotate the plot with text to state the time estimate alongside the curve

```{r}
median_est_2 <- median_est_1

median_est_2$plot <- median_est_2$plot +
  annotate("text", x = 450, y = 0.5, label = glue::glue(
    "The median survival \n time is ", 
    "{round(summary(fit1)$table['median'], 2)} days."), 
           col = msk_palette("contrast")[4], hjust = 0, fontface = 2)

ggsave("./img/median_est_2.png", print(median_est_2), width = 4, height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{median_est_2}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
We again use the \texttt{glue} function from the \texttt{glue} package to annotate our plot with reproducible text
  \end{column}
\end{columns}


## Inline R code can also be used to incorporate reproducible text into reports alongside graphical examples

\small

First define the `survfit` `summary` object, using the `survival` package:

\scriptsize

```{r echo = TRUE}
fit1 <- survival::survfit(survival::Surv(time, status) ~ 1, 
                          data = lung) 
res <- summary(fit1)$table
```

\small

Then:

The median survival time is `` `r
round(res["median"])` `` days (95% CI: `` `r
round(res["0.95LCL"])` `` - `` `r
round(res["0.95UCL"])` ``), which represents the estimated point in time that half of subjects will live beyond.

On knitting this will print our desired description:

The median survival time is `r round(res["median"])` days (95% CI: `r round(res["0.95LCL"])` - `r round(res["0.95UCL"])`), which represents the estimated point in time that half of subjects will live beyond.


## Finally put everything together into a report to our collaborator about the median survival time

```{r}
ggsave("./img/median_est_3.png", print(median_est_2), width = 5, height = 4.5)
```

\begin{center}
    \includegraphics[scale = 0.5]{median_est_3}
\end{center}

The median survival time is `r round(res["median"])` days (95% CI: `r round(res["0.95LCL"])` - `r round(res["0.95UCL"])`), which represents the estimated point in time that half of subjects will live beyond.


## Question from collaborator: Why can't I just estimate the median among those who had the event?

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{surv_curve_sq}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \begin{itemize}
  \setlength\itemsep{1em}
    \item[] In the \texttt{lung} data this would lead to an estimate of median survival time of `r summary(lung$time[lung$status == 2])["Median"]` days.
    \item[] But this is \textbf{incorrect}. 
    \item[] Why?
  \end{itemize}
  \end{column}
\end{columns}


## The true survival curve falls above a curve excluding the censored subjects

```{r message = FALSE}
fakedata <- lung %>% 
  filter(status == 2) %>% 
  mutate(group = "Ignoring censoring") %>% 
  full_join(mutate(lung, group = "With censoring"))

fit2 <- survfit(Surv(time, status) ~ group, data = fakedata)

wrong_median_1 <- ggsurvplot(data = fakedata, 
                         fit = fit2,
                         xlab = "Months",
                         palette = msk_palette("contrast")[c(3, 1)],
                         legend = "bottom",
                         legend.title = "",
                         legend.labs = c("Ignoring censoring", 
                                         "With censoring"),
                         xscale = 30.4,
                         break.x.by = 182.4, 
                         risk.table = TRUE, 
                         risk.table.y.text = FALSE)

ggsave("./img/wrong_median_1.png", print(wrong_median_1), width = 4.5, 
       height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_median_1}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
This graphical example compares the true survival curve in \textcolor{blue}{blue}, and the curve excluding censored subjects in \textcolor{purple}{purple}.
  \end{column}
\end{columns}


## Ignoring censoring leads to an underestimate of the median survival time

```{r}
wrong_median_2 <- wrong_median_1

wrong_median_2$plot <- wrong_median_2$plot  +
  geom_segment(x = -45, xend = 226, y = 0.5, yend = 0.5,
               col = msk_palette("contrast")[9], lwd = 1.2) +
  geom_segment(x = 226, xend = 226, y = 0.5, yend = -0.03, 
               col = msk_palette("contrast")[9], lwd = 1.2, 
               arrow = arrow(length = unit(0.1, "inches")))

ggsave("./img/wrong_median_2.png", print(wrong_median_2), width = 4.5, 
       height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_median_2}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
\begin{itemize}
  \setlength\itemsep{1em}
  \item[] The \textcolor{purple}{purple} group ignores censored subjects. 
  \item[] This is equivalent to the naive estimate of median time to death \textbf{among subjects who died} of `r summary(lung$time[lung$status == 2])["Median"]` days
  \end{itemize}
  \end{column}
\end{columns}


## Censored subjects contribute information because we know their event occurred after the censoring time

```{r}
wrong_median_3 <- wrong_median_2

wrong_median_3$plot <- wrong_median_2$plot + 
  geom_segment(x = -45, xend = 310, y = 0.5, yend = 0.5,
               col = msk_palette("main")[5], lwd = 1.2) +
  geom_segment(x = 310, xend = 310, y = 0.5, yend = -0.03, 
               col = msk_palette("main")[5], lwd = 1.2, 
               arrow = arrow(length = unit(0.1, "inches"))) +
  geom_segment(x = -45, xend = 226, y = 0.5, yend = 0.5,
               col = msk_palette("contrast")[9], lwd = 1.2, lty = 2) +
  geom_segment(x = 226, xend = 226, y = 0.5, yend = -0.03, 
               col = msk_palette("contrast")[9], lwd = 1.2, 
               arrow = arrow(length = unit(0.1, "inches"))) 

ggsave("./img/wrong_median_3.png", print(wrong_median_3), width = 4.5, 
       height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{wrong_median_3}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
\begin{itemize}
  \setlength\itemsep{1em}
    \item[] The \textcolor{blue}{blue} group includes the censored subjects. 
    \item[] The median survival time is longer, at $310$ days.
  \end{itemize}
  \end{column}
\end{columns}


## Question from collaborator: What does it mean when you say median survival is not reached?

```{r message = FALSE}
set.seed(20180902)
sample_data <- lung %>% 
  filter(status == 2) %>% 
  sample_frac(size = 0.15) %>% 
  full_join(lung %>% 
              filter(status == 1) %>% 
              sample_frac(size = 2, replace = T))
  
fit4 <- survfit(Surv(time, status) ~ 1, data = sample_data)

median_nr_1 <- ggsurvplot(data = lung, 
                         fit = fit4,
                         xlab = "Months",
                         palette = msk_palette("main")[3],
                         legend = "none",
                         xscale = 30.4,
                         break.x.by = 182.4, 
                        conf.int = FALSE, 
                        risk.table = TRUE,
                        risk.table.y.text = FALSE)

ggsave("./img/median_nr_1.png", print(median_nr_1), width = 4, height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{median_nr_1}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
\begin{itemize}
  \setlength\itemsep{1em}
    \item[] Survival curve based on a sample of \texttt{lung} data subjects
    \item[] In these data, we would report a median survival time of `r summary(fit4)$table["median"]`
  \end{itemize}
  \end{column}
\end{columns}


## The survival curve must cross the survival probability of $0.5$ before median survival time has been reached

```{r}
median_nr_2 <- median_nr_1

median_nr_2$plot <- median_nr_2$plot + 
  geom_segment(x = -45, xend = 1022, y = 0.5, yend = 0.5,
               col = msk_palette("contrast")[5], lwd = 1.2) +
  annotate("text", x = 90, y = 0.55, label = "Median", 
           col = msk_palette("contrast")[5], size = 5)

ggsave("./img/median_nr_2.png", print(median_nr_2), width = 4, height = 4.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.6]{median_nr_2}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
\begin{itemize}
  \setlength\itemsep{1em}
    \item[] By the end of the study, we have \textbf{not observed} the point in time that half the subjects will survive beyond
    \item[] Subjects would need to be followed longer, so that more deaths could occur, before we could observe median survival time
  \end{itemize}
  \end{column}
\end{columns}


## Question from collaborator: What is a hazard ratio?

```{r}
fit4 <- survfit(Surv(time, status) ~ sex, data = lung)

cph4 <- coxph(Surv(time, status) ~ sex, data = lung)

haz_plot <- ggsurvplot(data = lung, 
                       fit = fit4,
                       xlab = "Months",
                       palette = msk_palette("contrast")[6:7],
                       xscale = 30.4,
                       break.x.by = 182.4,
                       fun = "cumhaz",
                       legend.title = "",
                       legend.labs = c("Male", "Female"),
                       legend = "bottom", 
                       risk.table = TRUE,
                       risk.table.y.text = FALSE)

ggsave("./img/haz_plot.png", print(haz_plot), width = 4.5, height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{haz_plot}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
\begin{itemize}
  \setlength\itemsep{1em}
    \item[] The hazard ratio (HR) represents the ratio of hazards between two groups at any particular point in time
    \item[] The hazard of death for males versus females is `r round(exp(broom::tidy(cph4)["estimate"]), 2)` in the \texttt{lung} data
  \end{itemize}
  \end{column}
\end{columns}


## Hazard ratios apply at any given point in time, which is what differentiates them from relative risks

```{r}
haz_plot_2 <- haz_plot

haz_plot_2$plot <- haz_plot_2$plot + 
  annotate("text", x = 0, y = 3, label = glue::glue(
    "The hazard of death\nfor males versus females\n", 
    "is {round(exp(broom::tidy(cph4)['estimate']), 2)}."), 
    col = msk_palette("contrast")[12], size = 4, hjust = 0)
  
ggsave("./img/haz_plot_2.png", print(haz_plot_2), width = 4.5, height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{haz_plot_2}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
\begin{itemize}
  \setlength\itemsep{1em}
    \item[] HR = 0.5 would mean that at any particular time, half as many females as males are dying
    \item[] So our HR = `r round(exp(broom::tidy(cph4)['estimate']), 2)` implies that around 0.6 times as many females are dying as males, at any given time
  \end{itemize}
  \end{column}
\end{columns}


## Put it all together into an interpretable, reproducible report for your collaborator

```{r}
ggsave("./img/haz_plot_3.png", print(haz_plot_2), width = 5, height = 5.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
  \begin{center}
    \includegraphics[scale = 0.5]{haz_plot_2}
    \end{center}
  \end{column}
  
  \begin{column}{0.4\linewidth}
  \setlength\itemsep{1em}
The hazard ratio of `r round(exp(broom::tidy(cph4)['estimate']), 2)` implies that around 0.6 times as many females are dying as males, at any given time.
  \end{column}
\end{columns}


## Essential R packages 

\begin{itemize}
\item survival
\item survminer
\item ggplot2
\item glue
\item rmarkdown
\item knitr
\end{itemize}

\begin{center}
\includegraphics[scale=0.15]{hex}
\end{center}


## Thank you

Slides available at: \url{https://github.com/zabore/r-medicine}

\bigskip

\includegraphics[scale = 0.7]{contact}

