---
title: "Teaching Survival Analysis to Clinical Collaborators"
author: "Emily C. Zabor"
institute: "R/Medicine Conference"
date: "September 7, 2018"
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
library(tidyverse)
devtools::install_github("zabore/ezfun")
library(ezfun)
library(survival)
library(survminer)
library(knitr)
opts_chunk$set(echo = FALSE)
```


## The most common questions in cancer research relate to disease survival

\includegraphics[scale = 0.45]{pubs}

\begin{itemize}
  \item What is the probability of survival to a certain point in time?
  \item What is the average survival time?
\end{itemize}


## Survival analysis is a complex statistical procedure, so communication with collaborators is key

Strategies for conveying important information about survival analysis:

\begin{itemize}
  \setlength\itemsep{1em}
  \item[\checkbox] Be ready with \textbf{examples} to explain complex ideas
  \item[\checkbox] Use detailed \textbf{graphics} alongside the examples
  \item[\checkbox] Accompany numbers such as p-values and hazard ratios with detailed \textbf{explanations}
\end{itemize}


## Example of a dataset with censored data for a clinical application

The `lung` dataset is available from the `survival` package in `R`. The data contain subjects with advanced lung cancer from the North Central Cancer Treatment Group.

Variable descriptions, from the documentation:

\begin{itemize}
\item inst:	Institution code
\item \only<2>{\textcolor{blue}}{time:	Survival time in days}
\item \only<3>{\textcolor{blue}}{status:	censoring status 1=censored, 2=dead}
\item age: Age in years
\item sex: Male=1 Female=2
\item ph.ecog: ECOG performance score (0=good 5=dead)
\item ph.karno:	Karnofsky performance score (bad=0-good=100) rated by physician
\item pat.karno: Karnofsky performance score as rated by patient
\item meal.cal: Calories consumed at meals
\item wt.loss: Weight loss in last six months
\end{itemize}


## Censoring can be straightforward to understand in the context of a clinical trial

\begin{center}
\includegraphics[scale = 0.4]{trial_anatomy}
\end{center}

\bigskip

\tiny{RICH JT, NEELY JG, PANIELLO RC, VOELKER CCJ, NUSSENBAUM B, WANG EW. A PRACTICAL GUIDE TO UNDERSTANDING KAPLAN-MEIER CURVES. Otolaryngology head and neck surgery: official journal of American Academy of Otolaryngology Head and Neck Surgery. 2010;143(3):331-336. doi:10.1016/j.otohns.2010.05.007.}


## In retrospective data follow-up time is not fixed, but censoring still occurs

```{r message = FALSE}
hist_futime <- ggplot(lung, aes(x = time, fill = factor(status))) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = msk_palette("contrast")[c(4, 8)], 
                    labels = c("Censored", "Dead")) +
  theme_ezbasic() +
  labs(x = "Days",
       y = "Count")

ggsave("./img/hist_futime.png", hist_futime, width = 3.5, height = 3.5)
```

\begin{columns}
  \begin{column}{0.6\linewidth}
    \includegraphics[scale = 0.75]{hist_futime}
  \end{column}
  
  \begin{column}{0.4\linewidth}
    \begin{itemize}
      \setlength\itemsep{1em}
      \item Censored subjects had not yet died at date of data extraction
      \item Censored subjects still provide information
      \item The distribution of follow-up times is skewed
    \end{itemize}
  \end{column}
\end{columns}


## Question from collaborator: What is a survival curve?

```{r, message = FALSE}
fit1 <- survfit(Surv(time, status) ~ 1, data = lung)

surv_curve <- ggsurvplot(data = lung, 
                         fit = fit1,
                         xlab = "Months",
                         palette = msk_palette("main"),
                         legend = "none",
                         xscale = 30.4,
                         break.x.by = 182.4,
                         risk.table = TRUE,
                         risk.table.y.text = FALSE)

ggsave("./img/surv_curve.png", print(surv_curve), width = 5, height = 4.5)
```

\begin{center}
\includegraphics[scale = 0.55]{surv_curve}
\end{center}

The \textbf{Kaplan-Meier survival curve} is fundamental to survival analysis.


## The Kaplan-Meier survival function forms the basis of most survival analyses

Some basic facts:

\begin{itemize}
  \setlength\itemsep{1em}
  \item The x-axis is time and the y-axis is the survival function
  \item The survival function is calculated at each time as the ratio of subjects who did not experience the event by that time to the total number of subjects still at risk at that time
  \item Step function where each step down represents a time at which one or more events occurred
  \item Censored subjects are usually denoted by tick marks
\end{itemize}

  
## Question from collaborator: Can I report the percentage of events out of the total study population?

```{r, message = FALSE}
fakedata2 <- lung %>% 
  mutate(time = ifelse(status == 2, time, 1022), 
         group = "No censoring") %>% 
  full_join(mutate(lung, group = "With censoring"))

fit3 <- survfit(Surv(time, status) ~ group, data = fakedata2)

wrong_pct <- ggsurvplot(data = fakedata2, 
                         fit = fit3,
                         xlab = "Months",
                         palette = msk_palette("main")[c(2, 1)],
                         legend = "bottom",
                         legend.title = "",
                         legend.labs = c("No censoring", 
                                         "With censoring"),
                         xscale = 30.4,
                         break.x.by = 182.4)

ggsave("./img/wrong_pct.png", wrong_pct$plot, width = 5, height = 4)
```

\begin{center}
\includegraphics[scale = 0.55]{wrong_pct}
\end{center}

Imagine two studies, each with 228 subjects. There are 165 deaths in each study. The \textcolor{orange}{orange} study has no censoring during follow-up, the \textcolor{blue}{blue} study includes censored subjects (the true `lung` data).


## Ignoring censoring leads to an overestimate of the survival probability

\begin{itemize}
  \setlength\itemsep{1em}
  \item In the \textcolor{orange}{orange} study everyone is followed until the last follow-up time (i.e. no censoring during follow-up). $165/228$ subjects died during that time $\rightarrow$ survival probability at end of study = $1-\frac{165}{228} \times 100 = 28\%$.
    \item The \textcolor{blue}{blue} group has 63 subjects censored during follow-up. Still $165$ subjects died. 
    \item But the \textcolor{blue}{blue} curve is clearly lower than the \textcolor{orange}{orange} curve, especially at the end $\rightarrow$ survival probability at end of study = something $<28\%$ (actually $0.05\%$).
    \item This occurs because the censored subjects only contribute information for \textbf{part} of the follow-up time, and then they \textbf{fall out of the denominator}, thus pulling down the cumulative probability of survival.
  \end{itemize}


## Question from collaborator: What is the 1-year survival probability?

```{r}
est_1_year <- summary(fit1, times = 365.25)$surv

x_year_est <- ggsurvplot(data = lung, 
                         fit = fit1,
                         xlab = "Months",
                         palette = msk_palette("main"),
                         legend = "none",
                         risk.table = TRUE,
                         risk.table.y.text = FALSE,
                         xscale = 30.4,
                         break.x.by = 182.4)

x_year_est$plot <- x_year_est$plot + 
  geom_segment(x = 365.25, xend = 365.25, y = -0.05, yend = 0.4092416, 
               col = msk_palette("main")[4], size = 1.5) +
  geom_segment(x = 365.25, xend = -40, y = 0.4092416, yend = 0.4092416, 
               col = msk_palette("main")[4], size = 1.5, 
               arrow = arrow(length = unit(0.2, "inches"))) +
  annotate("text", x = -30, y = 0.25, 
           label = glue::glue("The 1-year survival \n probability is ", 
                              "{round(est_1_year, 2)}."), 
           col = msk_palette("main")[4], hjust = 0, fontface = 2)

ggsave("./img/x_year_est.png", print(x_year_est), width = 6, height = 4.5)
```

\begin{center}
\includegraphics[scale = 0.5]{x_year_est}
\end{center}

An annotated survival plot with reproducible text output using the `glue` package can help show that it is the probability on the y-axis corresponding to 1-year on the x-axis.


## Additionally include reproducible text with the image to make sure numbers are being correctly interpreted

As part of a reproducible workflow that includes writing reports in `R Markdown`, the `glue_data` function from the `glue` package can easily print reproducible text as a corresponding description:

\scriptsize

```{r echo = TRUE}
fit1 <- survival::survfit(
  survival::Surv(time, status) ~ 1, data = lung)
glue::glue_data(summary(fit1, times = 365.25), 
                "The survival probability at 1-year is ", 
                "{round(surv, 2)} ", 
                "(95% CI: ", "{round(lower, 2)} - ", "{round(upper, 2)}", 
                ")\n, which represents the proportion of ",
                "patients who survived beyond 1 year.")
```


## Question from collaborator: What is the median survival time?

```{r}
median_est <- ggsurvplot(data = lung, 
                         fit = fit1,
                         xlab = "Months",
                         palette = msk_palette("main"),
                         legend = "none",
                         risk.table = TRUE,
                         risk.table.y.text = FALSE,
                         xscale = 30.4,
                         break.x.by = 182.4)

median_est$plot <- median_est$plot + 
  geom_segment(x = -45, xend = 310, y = 0.5, yend = 0.5, 
               col = msk_palette("main")[4], size = 1.5) +
  geom_segment(x = 310, xend = 310, y = 0.5, yend = -0.03, 
               col = msk_palette("main")[4], size = 1.5, 
               arrow = arrow(length = unit(0.2, "inches"))) +
  annotate("text", x = -30, y = 0.25, label = glue::glue(
    "The median survival \n time is ", 
    "{round(summary(fit1)$table['median'], 2)} days."), 
           col = msk_palette("main")[4], hjust = 0, fontface = 2)

ggsave("./img/median_est.png", print(median_est), width = 6, height = 4.5)
```

\begin{center}
\includegraphics[scale = 0.5]{median_est}
\end{center}

The \textbf{median survival time} is the point on the x-axis corresponding to a survival probability of $0.5$ on the y-axis.


## Inline R code can also be used to incorporate reproducible text alongside graphical examples

\small

First define the `survfit` `summary` object:

\scriptsize

```{r echo = TRUE}
fit1 <- survfit(Surv(time, status) ~ 1, data = lung) 
res <- summary(fit1)$table
```

\small

Then:

The median survival time was `` `r
round(res["median"])` `` days (95% CI: `` `r
round(res["0.95LCL"])` `` - `` `r
round(res["0.95UCL"])` ``), which represents the point in time that half of subjects will live beyond.

On knitting this will print our desired description:

The median survival time was `r round(res["median"])` days (95% CI: `r round(res["0.95LCL"])` - `r round(res["0.95UCL"])`), which represents the point in time that half of subjects will live beyond.


## Question from collaborator: Why can't I just estimate the median among those who had the event?

```{r, message = FALSE}
fakedata <- lung %>% 
  filter(status == 2) %>% 
  mutate(group = "Ignoring censoring") %>% 
  full_join(mutate(lung, group = "With censoring"))

fit2 <- survfit(Surv(time, status) ~ group, data = fakedata)

wrong_median <- ggsurvplot(data = fakedata, 
                         fit = fit2,
                         xlab = "Months",
                         palette = msk_palette("contrast")[c(3, 1)],
                         legend = "bottom",
                         legend.title = "",
                         legend.labs = c("Ignoring censoring", 
                                         "With censoring"),
                         xscale = 30.4,
                         break.x.by = 182.4)

wrong_median$plot <- wrong_median$plot + 
  geom_segment(x = -45, xend = 310, y = 0.5, yend = 0.5,
               col = msk_palette("main")[5], lwd = 1.2) +
  geom_segment(x = 310, xend = 310, y = 0.5, yend = -0.03, 
               col = msk_palette("main")[5], lwd = 1.2, 
               arrow = arrow(length = unit(0.2, "inches"))) +
  geom_segment(x = -45, xend = 226, y = 0.5, yend = 0.5,
               col = msk_palette("contrast")[9], lwd = 1.2, lty = 2) +
  geom_segment(x = 226, xend = 226, y = 0.5, yend = -0.03, 
               col = msk_palette("contrast")[9], lwd = 1.2, 
               arrow = arrow(length = unit(0.2, "inches"))) +
  annotate("text", x = 300, y = 0.7, label = "Median survival time is\n underestimated when\n censoring is ignored.", 
           col = msk_palette("contrast")[9], hjust = 0, fontface = 2)

ggsave("./img/wrong_median.png", wrong_median$plot, width = 5, height = 4)
```

\begin{center}
\includegraphics[scale = 0.6]{wrong_median}
\end{center}

This graphical example compares the true survival curve in \textcolor{blue}{blue}, and the curve excluding censored subjects in \textcolor{purple}{purple}.


## Ignoring censoring leads to an underestimate of median survival time

\begin{itemize}
  \setlength\itemsep{1em}
  \item The \textcolor{purple}{purple} group ignores censored subjects. This is equivalent to a naive estimate of median time to death among subjects who died, which is $226$ days.
    \item The \textcolor{blue}{blue} group includes the censored subjects. The median survival time is clearly longer, at $310$ days.
    \item This occurs because the censored subjects do contribute information toward calculation of median survival time.
    \item Even though we don't know \textbf{exactly} when these patients died, we know they died \textbf{after} the time of censoring.
  \end{itemize}


## Question from collaborator: What does it mean when you say median survival is not reached?

```{r message = FALSE}
set.seed(20180902)
sample_data <- lung %>% 
  filter(status == 2) %>% 
  sample_frac(size = 0.15) %>% 
  full_join(lung %>% 
              filter(status == 1) %>% 
              sample_frac(size = 2, replace = T))
  
fit4 <- survfit(Surv(time, status) ~ 1, data = sample_data)

median_nr <- ggsurvplot(data = lung, 
                         fit = fit4,
                         xlab = "Months",
                         palette = msk_palette("main"),
                         legend = "none",
                         xscale = 30.4,
                         break.x.by = 182.4, 
                        conf.int = FALSE)

median_nr$plot <- median_nr$plot + 
  geom_segment(x = -45, xend = 1022, y = 0.5, yend = 0.5,
               col = msk_palette("main")[4], lwd = 1.2) +
  annotate("text", x = 45, y = 0.55, label = "Median", 
           col = msk_palette("main")[4], size = 5)

ggsave("./img/median_nr.png", median_nr$plot, width = 5, height = 3.5)
```

\begin{center}
\includegraphics[scale = 0.6]{median_nr}
\end{center}

The survival curve must \textbf{cross} the survival probability of $0.5$ before median survival time has been reached.

\scriptsize
(Survival curve based on a random sample of `lung` data subjects, sampling a proportion of those who died and oversampling censored subjects.)



## Median survial time is only observed when subjects are followed long enough 

When the median is not reached:

\begin{itemize}
\item By the end of the study, we have not observed the point in time that half the subjects will survive beyond
\item Subjects would need to be followed longer, so that more deaths could occur, before we could observe median survival time
\item Does \textbf{not} mean that the median survival time is greater than the maximum observed time in the study
\item Usually reported as "not reached" or "NR"
\end{itemize}


## Question from collaborator: What is a hazard ratio?

A \textbf{hazard ratio (HR)} represents the instantaneous chance of the event occurring in one group, divided by the instantaneous chance of the event occurring in the other group.

A simple example can help. We conduct a study comparing recurrence rates between treatment A and treatment B:

\begin{itemize}
\item \textbf{HR = 0.5}: at any particular time, \textbf{half} as many patients on treatment A are experiencing the event as compared to treatment B 
\item \textbf{HR = 1}: at any particular time, the \textbf{same} number of patients on treatment A are experiencing the event as compared to treatment B 
\item \textbf{HR = 2}: at any particular time, \textbf{twice} as many patients on treatment A are experiencing the event as compared to treatment B 
\end{itemize}


## Useful R packages 

\begin{itemize}
\item survival
\item survminer
\item ggplot2
\item glue
\item rmarkdown
\item knitr
\end{itemize}

\begin{center}
\includegraphics[scale=0.15]{hex}
\end{center}


## Thank you

Slides available at: \url{https://github.com/zabore/r-medicine}

\bigskip

\includegraphics[scale = 0.7]{contact}










```{r}
knit_exit()
```


## Question from collaborator: Can I report the percentage of events out of the total study population?

```{r}
basicdat <- data_frame(time = c(1, 2, 3, 4, 5, 5, 7, 8, 10, 10, 
                                1, 4, 5, 5, 8, 10, 10, 10, 10, 10),
                       event = c(1, 0, 0, 1, 1, 1, 0, 1, 1, 0, 
                                 1, 1, 1, 1, 1, 1, 0, 0, 0, 0), 
                       group = c(rep("censor", 10), rep("no_censor", 10)))

basicfit <- survfit(Surv(time, event) ~ group, data = basicdat)

basicplot <- ggsurvplot(data = basicdat,
           fit = basicfit, 
           palette = msk_palette("main"),
           legend = "bottom", 
           legend.title = "",
           legend.labs = c("Censoring", "No censoring"),
           conf.int = FALSE, 
           size = 1.2, 
           censor.size = 6)

ggsave("./img/basicplot.png", basicplot$plot, width = 5, height = 4)
```

\begin{center}
\includegraphics[scale = 0.55]{basicplot}
\end{center}

A simple graphical example can help. Imagine two studies, each with 10 subjects. There are 6 deaths in each group. The \textcolor{orange}{orange} group has no censoring, the \textcolor{blue}{blue} group has 3 subjects censored.


## Ignoring censoring leads to an overestimate of the survival probability

\begin{itemize}
  \setlength\itemsep{1em}
  \item The \textcolor{orange}{orange} group has no censoring. $6/10$ subjects die $\rightarrow$ survival probability at 10 years = $1-\frac{6}{10} \times 100 = 40\%$.
    \item The \textcolor{blue}{blue} group has 3 subjects censored. Still $6/10$ subjects die. 
    \item But the \textcolor{blue}{blue} curve is clearly lower than the \textcolor{orange}{orange} curve $\rightarrow$ survival probability at 10 years = something $<40\%$ (actually $17\%$).
    \item This occurs because the censored subjects only contribute information for \textbf{part} of the follow-up time.
  \end{itemize}


